---
title: "Survival Analysis Final Project"
author: Group 10
output: html_document
---

```{r message=FALSE, warning=FALSE}
library(readr)
library(tidyverse)
library(janitor)
library(ggcorrplot)
```

## EDA

### Load the dataset and clean column names

```{r}
file_path <- "Breast Cancer METABRIC.csv"

data <- read.csv(file_path, na.strings = "") %>%  # Treat blank strings as NA
  clean_names() %>%
  filter(cancer_type != "Breast Sarcoma") %>%  # only 3 patients are not Breast Cancer, exclude them
  rename(
    vital_status = patient_s_vital_status,
    three_gene_classifier_subtype = x3_gene_classifier_subtype
  ) %>%
  # Exclude patients with both overall_survival_months and overall_survival_status missing
  filter(!(is.na(overall_survival_months) & is.na(overall_survival_status)))

# View the cleaned data
head(data)
```

Our primary objective in this project is to investigate factors associated with overall survival (OS) in the METABRIC breast cancer cohort. Specifically, we define overall survival as the time from diagnosis to death from any cause, with overall_survival_months providing the follow-up time and overall_survival_status indicating whether the death event occurred.

For the survival analysis, both a valid survival time and an event indicator are required for each patient. In this dataset, some patients have overall_survival_months and overall_survival_status missing simultaneously, meaning that no follow-up or outcome information is available for these individuals. These cases are not censored observations but instead represent completely missing survival data. Therefore, we excluded patients with both overall_survival_months and overall_survival_status missing, so that the analysis dataset contains only patients with interpretable OS information for time-to-event modelling.

```{r}
dim(data)            # number of rows and columns
colnames(data)       # check all variable names
```

### Visualization of Overall Survival (OS)

```{r warning=FALSE}
ggplot(data, aes(x = overall_survival_months)) +
geom_histogram(bins = 20, color = "black", fill = "skyblue") +
labs(
title = "Distribution of Overall Survival Time",
x = "Overall survival time (months)",
y = "Count"
) +
theme_minimal()
```

```{r warning=FALSE}
ggplot(data, aes(x = factor(overall_survival_status))) +
  geom_bar(aes(y = ..count..), fill = "skyblue", color = "black") +  # Count bars
  geom_text(
    stat = "count",
    aes(label = paste0(..count.., " (", round(..count.. / sum(..count..) * 100, 1), "%)")),
    vjust = -0.3,      # move labels slightly above the bar
    color = "black",
    size = 4
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.1))  # add 10% headroom on top so labels are fully visible
  ) +
  labs(
    title = "Distribution of Overall Survival Status",
    x = "Overall Survival Status",
    y = "Count"
  ) +
  theme_minimal()
```

### Check for missing values:

```{r}
# Proportion and count of missing values per variable (only variables with missing values)
missing_summary <- data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%  # count missing values
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_count"
  ) %>%
  mutate(
    prop_missing = missing_count / nrow(data)  # calculate proportion of missing values
  ) %>%     # keep only variables with any missing values
  arrange(desc(missing_count))       # sort by missing count (descending)

missing_summary
```

```{r}
# Plot missing values per variable (count)
ggplot(missing_summary, aes(x = reorder(variable, missing_count), y = missing_count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Missing Values in Columns",
    x = "Variable",
    y = "Number of Missing Rows",
    caption = "Data: METABRIC Breast Cancer Study"
  ) +
  theme_minimal() +
  coord_flip()  # Flip the chart for better readability
```

### Define continuous and categorical variables manually:

#### Continuous variables:

```{r}
continuous_vars <- c(
  "age_at_diagnosis",
  "tumor_size",
  "lymph_nodes_examined_positive",
  "mutation_count",
  "nottingham_prognostic_index",
  "overall_survival_months",
  "relapse_free_status_months"
)

continuous_vars
```

```{r warning=FALSE}
# Histograms for all continuous variables
data %>%
  select(all_of(setdiff(continuous_vars, "overall_survival_months"))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(
    title = "Distributions of continuous variables",
    x = "Value",
    y = "Count"
  )
```

#### Categorical variables:

```{r}
# Treat other variables as categorical for EDA, but exclude patient_id
categorical_vars <- setdiff(names(data), c(continuous_vars, "patient_id"))
```

```{r}
# Frequency and proportion for each categorical variable (excluding overall_survival_status)
for (var in setdiff(categorical_vars, "overall_survival_status")) {
  cat("\n\n==============================\n")
  cat("Variable:", var, "\n")
  
  tmp <- data %>%
    count(.data[[var]]) %>%                 # count each level
    mutate(prop = n / sum(n)) %>%           # proportion within the variable
    arrange(desc(n))                        # sort by frequency
  
  print(tmp)                 # print all rows for this variable
}  
```

sex and cancer_type have only a single category (Female and Breast Cancer) in this dataset, making them uninformative for the survival analysis model, as they do not provide any variation in the data.

```{r}
# Remove 'sex' and 'cancer_type' from categorical_vars
categorical_vars <- setdiff(categorical_vars, c("sex", "cancer_type"))

# Print updated categorical_vars
categorical_vars
```

### Correlation analysis:

```{r warning=FALSE}
# Calculate correlation matrix for continuous variables
cor_matrix <- data %>%
  select(all_of(continuous_vars)) %>%
  cor(use = "pairwise.complete.obs")  # handle NA values by pairwise complete observation

ggcorrplot(cor_matrix, 
           type = "upper",        # Show only the lower triangle of the matrix
           lab = TRUE,            # Display correlation values
           colors = c("blue", "white", "red"),  # Color gradient
           title = "Correlation Matrix of Continuous Variables")
```

Based on the correlation matrix shown, relapse_free_status_months and overall_survival_months exhibit a very strong positive correlation of 0.9, indicating that these two variables are highly related. However, other continuous variables, such as nottingham_prognostic_index, mutation_count, lymph_nodes_examined_positive, and tumor_size, do not show high correlations with each other.

If we are conducting survival analysis with overall_survival_status as the event indicator, we should exclude relapse_free_status_months as a predictor in the model, as it is strongly correlated with overall_survival_months. Including both would lead to multicollinearity, where these two variables would essentially provide redundant information, making it difficult to assess their independent effects on survival. Therefore, we should choose overall_survival_months as the time variable, as it directly represents the total survival time and includes all relevant information related to the occurrence of the event.

```{r}
# Remove 'relapse_free_status_months' from continuous_vars
continuous_vars <- setdiff(continuous_vars, "relapse_free_status_months")
```

```{r warning=FALSE}
# Create an empty data frame to store results
chi_results <- data.frame(
  variable = character(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Perform Chi-Square Test for each categorical variable with overall_survival_status
for (var in categorical_vars) {
  
  # Create a contingency table for each variable and survival status
  contingency_table <- table(data[[var]], data$overall_survival_status)
  
  # Perform Chi-Square test
  chi_test <- chisq.test(contingency_table)
  
  # Store the results in chi_results dataframe
  chi_results <- rbind(chi_results, data.frame(variable = var, p_value = chi_test$p.value))
  
  # Print the contingency table and the p-value
  cat("\n\n==============================\n")
  cat("Variable:", var, "\n")
  print(contingency_table)
  cat("P-value for Chi-Square Test:", chi_test$p.value, "\n")
}

# Print the final results with p-values for all variables except overall_survival_status
chi_results
```

Since "vital_status" variable is already directly related to the outcome of survival and is a proxy for the status of the patient, it is unnecessary to include it in the model. This variable is highly correlated with survival, and adding it could lead to perfect prediction of the outcome or multicollinearity.

Although the chi-square test showed a statistically significant association between "cohort" and overall survival status, we decided not to include "cohort" as a covariate in the survival model. In this dataset, "cohort" primarily reflects study batch or recruitment group rather than a meaningful clinical or biological characteristic of the patients. Therefore, it is better interpreted as a design-related or administrative label, rather than a true risk factor for overall survival.

Although "relapse_free_status" shows a highly significant association with overall survival status in the chi-square test, we chose not to include it as a covariate in the overall survival model. Clinically, relapse is an intermediate outcome that occurs after baseline and lies on the causal pathway between baseline risk factors and death. Treating relapse status as a predictor of overall survival would therefore introduce information leakage and could distort the estimated effects of true baseline covariates. In addition, "relapse_free_status" is closely related to "relapse_free_status_months", which we already excluded as an alternative time-to-event endpoint.

```{r}
# Filter results to only show variables with p-value < 0.05 and exclude 'vital_status' and 'cohort'
significant_vars <- chi_results %>%
  filter(p_value < 0.05) %>%
  filter(!variable %in% c("vital_status","cohort","relapse_free_status"))

# Print the significant variables
significant_vars
```

### Final dataset after variable selection

```{r}
# Extract significant variables' names from significant_vars (filter for p-value < 0.05)
significant_var_names <- significant_vars$variable

# Filter the data by including only the selected variables from both significant_vars and continuous_vars
selected_vars <- c(significant_var_names, continuous_vars)

# Create a new dataset with only the selected variables
filtered_data <- data %>%
  select(all_of(selected_vars))

# View the filtered dataset
head(filtered_data)
```

### Overview of demographic and baseline variables

```{r}
#install.packages("gtsummary")
library(gtsummary)

vars_for_table1 <- c(
  "age_at_diagnosis",
  "tumor_size",
  "lymph_nodes_examined_positive",
  "mutation_count",
  "nottingham_prognostic_index",
  "type_of_breast_surgery",
  "cancer_type_detailed",
  "pam50_claudin_low_subtype",
  "neoplasm_histologic_grade",
  "tumor_other_histologic_subtype",
  "inferred_menopausal_state",
  "integrative_cluster",
  "radio_therapy",
  "three_gene_classifier_subtype",
  "tumor_stage"
)

table1 <-
  filtered_data |>
  dplyr::select(all_of(c("overall_survival_status", vars_for_table1))) |>
  tbl_summary(
    by = overall_survival_status,                
    type = list(all_continuous() ~ "continuous2"),
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",       
      all_categorical() ~ "{n} ({p}%)"           
    ),
    missing = "ifany"                           
  ) |>
  add_n() |>
  add_p() |>
  modify_caption("**Table 1. Baseline characteristics by overall survival status**") |>
  bold_labels()

table1
```
